---
title: "converting InterCatch exchange files"
author: "Edvin Fuglebakk"
date: "2024-03-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Convert ICEF to RCEF

An attempt at converting an InterCatch exchange file to a candidatr RCEF format.

## Format

Define variables available for population and domain definitions (in RCEF nomenclature)
```{r}
POPULATION_VARIABLES <- c("vesselFlagCountry", "year", "speciesCode", "catchCategory")
ACTIVITY_DOMAIN_VARIABLES <- c("quarter", "month", "fleet", "area")
FISH_DOMAIN_VARIABLES <- c("sex", "maturity", "minAge", "maxAge")
```

## Conversion

Define some general field-conversion functions

```{r}
convertCode <- function(code, codemap){
  if (all(code %in% names(codemap))){
    return(unlist(codemap[code]))
  }
  stop("Code not recognized.")
}

convertSpeciesCode <- function(FAOcode){
  codeMap <- list(POK=126441, COD=126436, HAD=126437, HER=126417, MAC=127023)
  return(convertCode(FAOcode, codeMap))
}
convertCatchCategory <- function(cgIC){
  codeMap <- list(L="Lan", D="Dis")
  return(convertCode(cgIC, codeMap))
}
convertUsage <- function(usageIC){
  codeMap <- list(H="Huc", I="Ind")
  return(convertCode(usageIC, codeMap))
}

convertMonth <- function(ICseasonType, ICseason){
  month <- ICseason*as.numeric(NA)
  month[ICseasonType=="Month"] <- ICseason[ICseasonType=="Month"]
  return(month)
}
convertQuarter <- function(ICseasonType, ICseason){
  quarter <- ICseason*as.numeric(NA)
  quarter[ICseasonType=="Month"] <- (ICseason[ICseasonType=="Month"] %/% 4) +1
  quarter[ICseasonType=="Quarter"] <- ICseason[ICseasonType=="Quarter"]
  return(quarter)
}
convertArea <- function(ICareaType, area){
  if (all(ICareaType %in% c("Div", "SubDiv", "SubArea", "Unit"))){
    return(area)
  }
  else{
    stop("Area not recognized.")
  }
}
#convert weight to kg per CL def
convertWeightKg <- function(unit, weight){
  factors <- rep(as.numeric(NA), length(unit))
  factors[unit=="kg"] <- 1
  factors[unit=="t"] <- 1e3
  return(weight*factors)
}

#' get activity domain from tables, after columns has been converted to RCEF.
getActvityDomain <- function(tab){
  tab <- data.table::as.data.table(tab)
  domainVars <- c(POPULATION_VARIABLES, ACTIVITY_DOMAIN_VARIABLES)
  return(apply(tab[,.SD,.SDcol=domainVars[domainVars %in% names(tab)]], 1, paste, collapse="/"))
}

convertSex <- function(ICsex){
 codeMap <- list('F'="F", 'M'="M", 'N'="U")
  return(convertCode(ICsex, codeMap)) 
}

convertMaturity <- function(ICmaturity){
  if (all(is.na(ICmaturity))){
    return(ICmaturity)
  }
  stop("maturity conversion not implemented.")
 codeMap <- list()
  return(convertCode(ICmaturity, codeMap)) 
}

#' get fish domain from tables, after columns has been converted to RCEF.
getFishDomain <- function(tab){
  tab <- data.table::as.data.table(tab)
  domainVars <- c(POPULATION_VARIABLES, FISH_DOMAIN_VARIABLES)
  return(apply(tab[,.SD,.SDcol=domainVars[domainVars %in% names(tab)]], 1, paste, collapse="/"))
}

convertNumber <- function(unit, number){
  factor <- rep(as.numeric(NA), length(number))
  factor[unit=="k"] <- 1e3
  factor[unit=="m"] <- 1e6
  factor[unit=="n"] <- 1
  return(number*factor)
}

#convert weight to g per definition for weightlive (https://vocab.ices.dk/?ref=1606)
convertWeightG <- function(unit, weight){
  factors <- rep(as.numeric(NA), length(unit))
  factors[unit=="g"] <- 1
  factors[unit=="kg"] <- 1e3
  return(weight*factors)
}

#tentatively convert length without specifying measurment
convertLengthCm <- function(unit, lengthCm){
  factors <- rep(as.numeric(NA), length(unit))
  factors[unit=="cm"] <- 1
  factors[unit=="mm"] <- 1e-1
  return(lengthCm*factors)
}

convertFishingDays <- function(unit, effort){
  eff <- rep(as.numeric(NA), length(unit))
  eff[unit=="fd"] <- effort[unit=="fd"]
  return(eff)
}

```


Define functions to extract fish domains. They need to be extracted both for making the FISHDOMAIN table and the ESTIMATE table of RCEF, so we define an auxilirary function that supports both. In order to make the fish domains, we need to specify a max age for the plus group, which is not defined in InterCatch:

```{r}
#' extract fish domain portion of SDtable, keep all rows in same order as SDtabe.
extractSDWfishDomains <- function(SDtable, plusgroupMax){
  fishdomain <- SDtable
  
  #population variables
  fishdomain <- SDtable[,c("Country", "Year")]
  names(fishdomain) <- c("vesselFlagCountry", "year")
  fishdomain$speciesCode <- convertSpeciesCode(SDtable$Species)
  fishdomain$catchCategory <- convertCatchCategory(SDtable$CatchCategory)
  
  
  fishdomain$sex <- convertSex(SDtable$Sex)
  fishdomain$maturity <- convertMaturity(SDtable$Maturity)
  
  if (any(SDtable$CANUMtype=="Age")){
    fishdomain$minAge <- as.numeric(NA)
    fishdomain$maxAge <- as.numeric(NA)
    fishdomain$minAge[SDtable$CANUMtype=="Age"] <- SDtable$AgeLength[SDtable$CANUMtype=="Age"]
    fishdomain$maxAge[SDtable$CANUMtype=="Age"] <- SDtable$AgeLength[SDtable$CANUMtype=="Age"]
    fishdomain$maxAge[SDtable$CANUMtype=="Age" & SDtable$AgeLength==SDtable$PlusGroup] <- plusgroupMax
  }
  
  if (any(SDtable$CANUMtype=="Length")){
    stop("Conversion for length not supported.")
  }
  
  for (col in names(fishdomain)){
    if (all(is.na(fishdomain[[col]]))){
      fishdomain[[col]] <- NULL
    }
  }

  fishdomain$fishDomain <- getFishDomain(fishdomain)
  return(fishdomain)
}

#' define fish domains
#' @param SDtable
#' @param plusgroupMax the value to use as max age for any group representing an IC plusgroup
extractFishDomain <- function(SDtable, plusgroupMax){
  fishdomain <- extractSDWfishDomains(SDtable, plusgroupMax)
  fishdomain <- fishdomain[!duplicated(fishdomain$fishDomain),]
  return(fishdomain)
}
```

Define functions to extract activity domains. They need to be extracted for making the CATCH table and the ESTIMATE table of RCEF, so we define an auxilirary function that supports both, and we will define the function to extract the CATCH table:

```{r}
#' extract population variables and activity domain variables, and annotate activity domain
extractActivityDomains <- function(table){
  ACTDOM <- table[,c("Country", "Year")]
  names(ACTDOM) <- c("vesselFlagCountry", "year")
  ACTDOM$speciesCode <- convertSpeciesCode(table$Species)
  ACTDOM$catchCategory <- convertCatchCategory(table$CatchCategory)
  
  ACTDOM$quarter <- convertQuarter(table$SeasonType, table$Season)
  ACTDOM$month <- convertMonth(table$SeasonType, table$Season)
  ACTDOM$fleet <- table$Fleet
  ACTDOM$area <- convertArea(table$AreaType, table$FishingArea)
  if (!all(is.na(table$DepthRange))){
    stop("Support for the field DepthRange not implemented.")
  }
  if (!all(is.na(table$Stock))){
    stop("Support for the field Stock not implemented.")
  }
  if (!all(table$ReportingCategory=="R")){
    stop("Conversion only defined for reporting category 'R'")
  }
  if (!all(is.na(table$DataToFrom))){
    stop("Conversion only defined for reporting category 'R', DataToFrom should be NA.")
  }
  if (!all(is.na(table$QualityFlag))){
    stop("Support for the field QualityFlag is not implemented.")
  }
  
  ACTDOM$activityDomain <- getActvityDomain(ACTDOM)
  
  return(ACTDOM)
}

extractCatch <- function(SItable){
  
  #population variables
  CATCH <- extractActivityDomains(SItable)
  if (any(CATCH$catchCategory!="Lan")){
    stop("Only support for catchCategory 'Lan' is implemented")
  }
  
  CATCH$landingCategory <- convertUsage(SItable$Usage)
  
  CATCH$scientificWeight <- convertWeightKg(SItable$UnitCATON, SItable$CATON)
  CATCH$officialWeight <- convertWeightKg(SItable$UnitCATON, SItable$Offlandings)
  if (!all(is.na(SItable$varCATON))){
    stop("Support for the field varCATON is not implemented.")
  }
  if ("infoFleet" %in% names(SItable)){
    CATCH$infoFLeet <- SItable$infoFleet
    if (all(nchar(CATCH$infoFleet)==0)){
      CATCH$infoFleet <- as.character(NA)
    }
  }
  if ("InfoStockCoordinator" %in% names(SItable)){
    CATCH$InfoStockCoordinator <- SItable$InfoStockCoordinator    
    if (all(nchar(CATCH$InfoStockCoordinator)==0)){
      CATCH$InfoStockCoordinator <- as.character(NA)
    }
  }
  if ("InfoGeneral" %in% names(SItable)){
    CATCH$InfoGeneral <- SItable$InfoGeneral
    if (all(nchar(CATCH$InfoGeneral)==0)){
      CATCH$InfoGeneral <- as.character(NA)
    }
  }
  
  for (col in names(CATCH)){
    if (all(is.na(CATCH[[col]]))){
      CATCH[[col]] <- NULL
    }
  }

  return(CATCH)
  
}
```

We will define a function to extract the EFFORT table:

```{r}
extractEffort <- function(SItable){  
  effort <- SItable[,c("Country", "Year")]
  names(effort) <- c("vesselFlagCountry", "year")
  
  effort$quarter <- convertQuarter(SItable$SeasonType, SItable$Season)
  effort$month <- convertMonth(SItable$SeasonType, SItable$Season)
  effort$fleet <- SItable$Fleet
  effort$area <- convertArea(SItable$AreaType, SItable$FishingArea)
  if (!all(is.na(SItable$DepthRange))){
    stop("Support for the field DepthRange not implemented.")
  }
  if (!all(is.na(SItable$AreaQualifier))){
    stop("Support for the field AreaQualifier not implemented.")
  }
  if (!all(is.na(SItable$unitEffort) | (SItable$unitEffort %in% c("fd")))){
    stop("Support the some effort units are not implemented.")
  }
  effort$officalFishingDays <- convertFishingDays(effort$unitEffort, effort$effort)
  
  return(effort)
}
```

We will define a function to extract the ESTIMATE table. Mostly this is constructed from the SDtable, but it also need access to one variable at the SI table. In addtion it needs to be informed about two parameters not encoded in InterCatch:

```{r}
extractEstimates <- function(SDtable, SItable, plusgroupMax, lengthMeasurement){
  
  # look up samples origin for each domain
  SIactDom <- extractActivityDomains(SItable)
  SIactDom$SamplesOrigin <- SItable$SamplesOrigin
  SIactDom <- SIactDom[!duplicated(SIactDom$activityDomain),c("activityDomain", "SamplesOrigin")]
  
  #annotate domains
  estimates <- extractSDWfishDomains(SDtable, plusgroupMax)
  activityDomains <- extractActivityDomains(SDtable)
  estimates <- cbind(estimates, activityDomains[,.SD, .SDcol=names(activityDomains)[!(names(activityDomains) %in% POPULATION_VARIABLES)]])
  
  estimates <- estimates[,.SD, .SDcol=c(POPULATION_VARIABLES, "activityDomain", "fishDomain")]
  
  #add values
  estimates$totalNumber <- convertNumber(SDtable$unitCANUM, SDtable$NumberCaught)
  estimates$meanWeightLive <- convertWeightG(SDtable$unitMeanWeight, SDtable$MeanWeight)
  
  if (any(!is.na(SDtable$MeanLength)) & lengthMeasurement != "TotalLength"){
    stop("Only conversion with total length is implemented.")
  }

  estimates$meanTotalLength <- convertLengthCm(SDtable$UnitMeanLength, SDtable$MeanLength)
  
  #add sample descriptions
  estimates$samplesOrigin <- SIactDom$SamplesOrigin[match(estimates$activityDomain, SIactDom$activityDomain)]
  estimates$sampledCatch <- SDtable$SampledCatch
  estimates$numSamplesLngt <- SDtable$NumSamplesLngt
  estimates$numLngtMeas <- SDtable$NumLngtMeas  
  estimates$numSamplesAge <- SDtable$NumSamplesAge  
  estimates$numAgeMeas <- SDtable$NumAgeMeas  
  return(estimates)
}
```

We combine these to make a "complete" conversion function:

```{r}
#' Convert IC data to RCEF
#' @param IC Data in InterCatch Exchange format, as parsed by RstoxData::parseInterCatch
#' @param plusgroupMax the value to use as max age for any group representing an IC plusgroup
convert <- function(IC, plusgroupMax=100, lengthMeasurement="TotalLength"){
  RCEF <- list()
  RCEF$CATCH <- extractCatch(IC$SI)
  RCEF$FISHDOMAIN <- extractFishDomain(IC$SD, plusgroupMax)
  RCEF$ESTIMATES <- extractEstimates(IC$SD, IC$SI, plusgroupMax, lengthMeasurement)
  RCEF$EFFORT <- extractEffort(IC$SI)
  
  return(RCEF)
}
```

And apply it to one example file:

```{r}
ICexample <- RstoxData::parseInterCatch("../data/intercatch_example.csv")
RCEF <- convert(ICexample)
```


## Notes and Issues

### Usage
It is a bit unclear if usage is (can be) an activity variable when converting from InterCatch. It does not exist on the SD-table. Unsure if it is possible to have SI records that differ in Usage, but is equal for all the mandatory fields.

### Length
Did not test conversion for length groups. Is IC agnostic about which length measurement is used? If so, we will have to convert with parameterization, as is tentatively done for meanTotalLength.

### Weight
I think it is correct to map CATON from InterCatch to scientificWeight in CL, and OffLandings in InterCatch to officialWeight in CL. Need to check this.

### Area
IC uses explicit declaration of the kind of area. CL uses "most detailed FAO area". If this is taken to mean "most detailed available in data", we can convert from IC as above. If it is taken to mean "FAO area with the highest precision defined" we may have to include one column for SubArea, one for Division, one for SubDivision and so on.

### Capitalization and naming
I have spelled fields names in RCEF without capitalization. In the example files we have used a mix. Should decide on something.

## Columns that need to be added
The conversion excercies revealed some columns that need to be added to the RCEF draft.

### Columns needed from CL (not in draft)
landingCategory
scientificWeight

### Columns needed from BiologicalMeasurments (not in draft)
maturity

### Columns needed from IC (not in draft)
SamplesOrigin
InfoFleet
InfoStockCoordinator
InfoGeneral

### Columns not included, with unclear status in IC (not in draft)
DepthRange
Stock
QualityFlag
varCATON
varNumLanded
varWgtLanded
varLgtLanded

### Columns already included from CL (already in draft)
vesselFlagCountry
speciesCode
catchCategory
quarter
month
officialWeight

### Columns already included from BiologicalMeasurments (already in draft)
sex

### Columns already included from IC (already in draft)
fleet
SampledCatch
NumSamplesLngt
NumLngtMeas
NumSamplesAge
NumAgeMeas