---
title: "RCEF stock coordination"
author: "Edvin Fuglebakk"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Some proof-of-concept checks to verify that candidates for an exchange format for National / Regional estimation in RDBES/TAF can be used for stock-coordination purposes.

## Format
We will define some preliminary vocabularies for the columns that are valid in the RCEF format:

```{r format}
POPULATION_VARIABLES <- c("vesselFlagCountry", "year", "catchCategory", "speciesCode")
ACTIVITY_DOMAIN_VARIABLES <- c("Quarter",	"Month","stockArea",	"area",	"metier6",	"Fleet")
FISH_DOMAIN_VARIABLES <- c("MinAge",	"MaxAge", "Sex")
ESTIMATES_VALUE_VARIABLES <- c("totalNumber", "totalWeightLive", "varTotalNumber", "varTotalWeight")
SAMPLE_DESCRIPTION_VARIABLES <- c("numSamplesLength", "numLngMeas", "numSamplesAge", "sampledCatch")
```

Some example files are located in ../data .

## Aggregate domains

An example of how estimates for activity domains can be aggregated is provided in the following function:

```{r aggregateEstimates}
#' @param RCEF domain definitionns and estimates in the RCEF format.
#' @param desiredDomains table defining desiredDomains. Contain valid domain variables and a column 'activityDomain'
aggregateActivityDomains <- function(RCEF, desiredDomains){

  catch <- data.table::as.data.table(RCEF$CATCH)
  estimates <- data.table::as.data.table(RCEF$ESTIMATES)
  desiredDomains <- data.table::as.data.table(desiredDomains)
  
  #check that the desired domain is defined in terms of valid activity domain variables  
  stopifnot(all(names(desiredDomains) %in% c(ACTIVITY_DOMAIN_VARIABLES, "activityDomain")))
  
  #check that all activity domains in estimates are defined in the catch table
  stopifnot(all(catch$activityDomain %in% catch$activityDomain))
  
  #handle domain variables not in domain definition for estimates
  domainCount <- desiredDomains[,list(NDomains=length(unique(activityDomain))), by=names(desiredDomains)[!(names(desiredDomains) %in% names(catch))]]
  if (ncol(domainCount)>1 & any(domainCount$NDomains>1)){
    stop("Desired domains specify finer resolution that domains estimates are provided for")
  }
  
  # domains encompass all values for domain variables not provided
  desiredDomains <- desiredDomains[,.SD, .SDcol=names(desiredDomains)[names(desiredDomains) %in% names(catch)]]
  if (ncol(desiredDomains) == 1){
    soptifnot(length(unique(desiredDomain$activtyDomain))!=1)
    catch$actvityDomain <- desiredDomain$activtyDomain
  }
  else{
    catch <- merge(catch, desiredDomains, by=names(desiredDomains)[names(desiredDomains)!="activityDomain"], suffixes=c("", ".desired"))
    
    domainCount <- catch[,list(NDomains=unique(activityDomain.desired)), by="activityDomain"]
    if (any(domainCount$Ndomains>1)){
      stop("Some estimated domains are in more than one desired domain.")
    }
  }
  
  estimates$activityDomain.desired <- catch$activityDomain.desired[match(estimates$activityDomain, catch$activityDomain)]
  
  catch$activityDomain <- catch$activityDomain.desired
  catch$activityDomain.desired <- NULL
  
  estimates$activityDomain <- estimates$activityDomain.desired
  estimates$activityDomain.desired <- NULL
  
  aggregated <- NULL
  # iterate over value variables and add
  for (e in names(estimates)){
    if (e %in% ESTIMATES_VALUE_VARIABLES){
      agg <- estimates[,list(agg=sum(get(e))), by=c("fishDomain", "activityDomain", POPULATION_VARIABLES)]
      names(agg)[names(agg)=="agg"] <- e
      
      if (is.null(aggregated)){
        aggregated <- agg
      }
      else{
        aggregated <- merge(aggregated, agg)
      }
    }
  }
  # iterate over sample variables and set to NA
  for (e in names(estimates)){
    if (e %in% SAMPLE_DESCRIPTION_VARIABLES){
      aggregated[[e]] <- as.numeric(NA)
    }
  }
  # Compute aggregated mean weight
  agg <- estimates[,list(meanWeightLive = sum(meanWeightLive*totalNumber)/sum(totalNumber), varMeanWeightLive = sum(varMeanWeightLive * (totalNumber/sum(totalNumber))**2)), by=c("fishDomain", "activityDomain", POPULATION_VARIABLES)]
  if (is.null(aggregated)){
        aggregated <- agg
  }
  else{
        aggregated <- merge(aggregated, agg)
  }

  RCEF$ESTIMATES <- aggregated
  RCEF$CATCH <- catch
  
  return(RCEF)
}
```

Note that aggregation of means requires different treatment than the simply additive statistics. May consider a more general encoding, where we parameterise variable, and provide Total, Mean, VarTotal, and VarMean for that variable. That was also one of the proposed solutions to handle different length variables (total length, carapax length, fork length, etc).

It is not clear that sample description parameters can be aggregated across activity domains. If PSU is for instance a landing event, some activity domain variables like gear can be covered by the same samples and it is unclear if it can be added. Perhaps these should just be NA as here, be defined as lower bound on sampling, or we need a slightly different format for aggregated results omitting them altogether.

Applying this function to an example file with the estimates reported by age and gear:

```{r aggregateGear}
domainEstimates <- readRDS("../data/estimates_gear_caa.rds")
desiredDomains <- data.table::data.table(activityDomain="NSHER", metier6=c("OTB_SPF_>0_0_0_all","OTM_SPF_>0_0_0_all","PS1_SPF_>0_0_0_all"))
totalAggregatedEstimates <- aggregateActivityDomains(domainEstimates, desiredDomains)
totalAggregatedEstimates$ESTIMATES
```
Similarly, grouping only two gears:

```{r aggregateGear2}
desiredDomains <- data.table::data.table(activityDomain=c("OT", "OT", "PS"), metier6=c("OTB_SPF_>0_0_0_all","OTM_SPF_>0_0_0_all","PS1_SPF_>0_0_0_all"))
aggregatedEstimates <- aggregateActivityDomains(domainEstimates, desiredDomains)
aggregatedEstimates$ESTIMATES
```


## Imputing mean weight

If estimates cannot be provided for a certain activity domain, or they are provided with too low precision, one may desire to impute mean weight and weigth fraction in fishdomains (e.g. age groups) from a different activity domain, and provide totalNumber as a ratio estimate. An example is provided in the following function:

```{r imputeMeanWeight}
#' @param RCEF
#' @param imputeTo The activity domain to impute to
#' @param imputeFrom The activity domain to impute from
imputeMeanWeight <- function(RCEF, imputeTo, imputeFrom){
  
  stopifnot(imputeFrom %in% RCEF$CATCH$activityDomain)
  stopifnot(imputeFrom %in% RCEF$ESTIMATES$activityDomain)
  stopifnot(imputeTo %in% RCEF$CATCH$activityDomain)

  estimatesImputed <- RCEF$ESTIMATES[activityDomain==imputeFrom,]
  estimatesImputed$fraction <- estimatesImputed$totalNumber * estimatesImputed$meanWeightLive / sum(estimatesImputed$totalNumber * estimatesImputed$meanWeightLive)
  estimatesImputed$varFraction <- estimatesImputed$varTotalNumber * (estimatesImputed$meanWeightLive / sum(estimatesImputed$totalNumber * estimatesImputed$meanWeightLive))**2
  
  for (e in names(estimatesImputed)){
    if (e %in% ESTIMATES_VALUE_VARIABLES){
      estimatesImputed[[e]] <- NA
    }
  }
  for (e in names(estimatesImputed)){
    if ( e %in% SAMPLE_DESCRIPTION_VARIABLES){
    estimatesImputed[[e]] <- 0      
    }
  }
  estimatesImputed$activityDomain <- imputeTo
  totalWeight <- sum(RCEF$CATCH$officialWeight[RCEF$CATCH$activityDomain == imputeTo])*1000
  
  estimatesImputed$totalNumber <- totalWeight * estimatesImputed$fraction / estimatesImputed$meanWeightLive
  estimatesImputed$varTotalNumber <- estimatesImputed$varFraction * (totalWeight / estimatesImputed$meanWeightLive)**2
  estimatesImputed$fraction <- NULL
  estimatesImputed$varFraction <- NULL
  
  estimates <- RCEF$ESTIMATES[RCEF$ESTIMATES$activityDomain != imputeTo,]
  estimates <- rbind(estimates, estimatesImputed)
  
  RCEF$ESTIMATES <- estimates
  return(RCEF)
}
```

Note that in this case, it is quite useful to be able to explicitly set the sampling description parameters to 0. Perhaps supporting the idea of letting these variables be a lower bound.

Applying this to the example file that is missing estimate for a domain, and aggregating:

```{r imputeByGear}
#impute for missing domains
partialDomainEstimates <- readRDS("../data/estimates_gear_caa_missing_domains.rds")
imputedDomainEstimates <- imputeMeanWeight(partialDomainEstimates, "NSHER/OTB", "NSHER/OTM")
imputedDomainEstimates <- imputeMeanWeight(imputedDomainEstimates, "NSHER/PS1", "NSHER/OTM")

#add up to grand total
desiredDomains <- data.table::data.table(activityDomain="NSHER", metier6=c("OTB_SPF_>0_0_0_all","OTM_SPF_>0_0_0_all","PS1_SPF_>0_0_0_all"))
imputedTotalEstimates <- aggregateActivityDomains(imputedDomainEstimates, desiredDomains)

imputedDomainEstimates$ESTIMATES
imputedTotalEstimates$ESTIMATES
```

## Comparisons

As a quick quality check we compare the aggregated estimated and the aggregated imputed estimate with the direct estimate of total number in each age group (provided in ../data):

```{r comparisons}
totalEstimates <- readRDS("../data/estimates_caa.rds")
totalEstimates$ESTIMATES$Method="Direct"
imputedTotalEstimates$ESTIMATES$Method="Imputed"
totalAggregatedEstimates$ESTIMATES$Method="Aggregated"

comparison <- rbind(totalEstimates$ESTIMATES, imputedTotalEstimates$ESTIMATES, totalAggregatedEstimates$ESTIMATES)

ggplot2::ggplot(comparison) +
    ggplot2::geom_col(ggplot2::aes(x=fishDomain, y=totalNumber, fill=Method), position="dodge") +
    ggplot2::geom_errorbar(ggplot2::aes(x=fishDomain, ymin=totalNumber - sqrt(varTotalNumber), ymax=totalNumber + sqrt(varTotalNumber), group=Method), width=.2, position=ggplot2::position_dodge(.9)) +
  ggplot2::ylab("total number +/- SE") +
  ggplot2::xlab("Age group")

```

## Issues

* Should we change the representation of values in the ESTIMATES table, so that more generic aggregation scripts can be made? For instance represent all variables as VARIABLE, TOTAL, MEAN, TOTALVAR, and MEANVAR. In addition to TOTALNUMBER and TOTALNUMBERVAR. This would also provide a solution to the open issue about how to represent different length measurements (total length, carapax length, etc.)
* How should we deal with sample description parameters. It seems they cannot generally be aggregated across domains. Redifining them to be a lower bound makes it trivial to aggregate, and allows us to label imputed domains with sample size 0. Otherwise they are probably not important to keep after stock-coordination, and we could set them to NA. Having stock-coordination routines return the same format as they accept, facilitates stringing together stock-coordination, as is done repeatedly above.
